---
title: "Session 4 Exercise: Diagnosing a Wage Regression"
subtitle: "Quantitative Data Science · Europa-Universität Flensburg"
author: "Your name here"
date: "2026-05-21"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-tools: true
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

::: {.callout-note}
**Business question:** *What determines hourly wages in the US labour market, and when can we trust a regression-based answer?*

Work through the tasks below in order. For each task, write your R code in the provided chunk and your interpretation directly below it. Aim for complete sentences — imagine you are explaining your findings to a colleague who has not studied econometrics.

You have approximately **50 minutes** for the core tasks.
:::

---

# Setup

```{r}
library(tidyverse)
library(broom)
library(modelsummary)
library(lmtest)
library(sandwich)
library(car)
library(performance)
library(see)
library(patchwork)

cps <- read_csv("data/cps-earnings-2014.csv", show_col_types = FALSE)
```

The dataset is a cleaned extract from the 2014 US Current Population Survey (CPS), restricted to full-time employed workers (≥ 35 hours/week). Variables:

| Variable | Description |
|---|---|
| `wage` | Hourly wage (USD) = weekly earnings / usual weekly hours |
| `earnwke` | Weekly earnings (USD) — top-coded at $2,884 |
| `uhours` | Usual hours worked per week |
| `age` | Age in years (16–64) |
| `educ` | Educational attainment (CPS grade92 codes: 31 = < grade 1, 46 = doctoral degree) |
| `female` | 1 = female, 0 = male |

Take two minutes to look at the data before modelling:

```{r}
glimpse(cps)
summary(cps[, c("wage", "age", "educ", "female")])
```

**Question:** Look at the wage distribution. Is it symmetric or skewed? What does this suggest about the appropriate functional form?

*Your answer:*

---

# Task 1 — Functional form {#task-1}

A classic result in labour economics (the "Mincer equation") is that log wages — not wages in levels — tend to follow an approximately linear relationship with education and age.

**1a.** Estimate two specifications and extract residuals with `broom::augment()`:

```{r}
m_level <- lm(wage       ~ age + educ + female, data = cps)
m_log   <- lm(log(wage)  ~ age + educ + female, data = cps)
```

Build a **residuals-vs-fitted plot** for each model. Use `ggplot2` and a loess smoother. What does the smoother tell you about each specification?

```{r}
# your code
```

*Your answer:*

**1b.** Run the RESET test on both models. What do you conclude — and what does a significant result tell you (and not tell you)?

```{r}
# your code — resettest(m_level), resettest(m_log)
```

*Your answer:*

**1c.** Adopt `m_log` as your working model. Interpret the coefficient on `educ` in one sentence. (Recall: `educ` is the CPS ordinal education code, not years of schooling.)

*Your answer:*

---

# Task 2 — Heteroskedasticity {#task-2}

**2a.** Build a **scale-location plot** for `m_log` manually: plot $\sqrt{|\tilde{e}_i|}$ against fitted values. A flat horizontal band indicates constant variance.

```{r}
# augment(m_log) |> ggplot(...) + geom_smooth(...)
```

*Your answer:*

**2b.** Run the Breusch-Pagan test.

```{r}
# your code — bptest(m_log)
```

**2c.** Compare `m_log` with classical and HC1 robust standard errors in a `modelsummary()` table.

```{r}
# modelsummary(list("Classical" = m_log, "HC1 robust" = m_log), vcov = list(...))
```

**Questions:**
- Does wage data show heteroskedasticity? What is the economic intuition for why you would expect this?
- Which coefficients change significance when you switch to robust SEs?
- Should you always use robust SEs? Write one sentence defending your position.

*Your answer:*

---

# Task 3 — Multicollinearity {#task-3}

**3a.** Compute VIF for `m_log`.

```{r}
# your code — vif(m_log)
```

**3b.** Estimate an extended model adding `I(age^2)` (to allow a non-linear age-wage profile) and re-compute VIF.

```{r}
m_quad <- lm(log(wage) ~ age + I(age^2) + educ + female, data = cps)
# vif(m_quad)
```

**Questions:**
- Are VIF values for `m_log` a concern?
- `age` and `I(age^2)` will have high VIF in `m_quad`. Is this a problem? Why or why not?
- Would you use `m_log` or `m_quad` as your working model? Look at the residual plot for both specifications.

```{r}
# Compare residual plots for m_log and m_quad
```

*Your answer:*

---

# Task 4 — Influential observations {#task-4}

**4a.** Use `augment()` to build a **leverage vs. standardised residuals plot** with Cook's distance contours. Use the $4/n$ threshold to colour influential observations.

```{r}
threshold  <- 4 / nrow(cps)
p_params   <- length(coef(m_log))
cps_aug    <- augment(m_log)

# Build h_seq for contour curves:
# h_seq <- seq(1e-5, max(cps_aug$.hat) * 1.05, length.out = 500)
# contour_df <- tibble(h = h_seq,
#                      upper =  sqrt(threshold * p_params * (1 - h_seq) / h_seq),
#                      lower = -sqrt(threshold * p_params * (1 - h_seq) / h_seq))
# your plot code
```

**4b.** List the ten most influential workers. What do they have in common?

```{r}
# cps_aug |> arrange(desc(.cooksd)) |> head(10) |> ...
```

**4c.** Re-estimate `m_log` excluding influential observations. Does the main conclusion about education and wages change?

```{r}
# your code
```

*Your answer:*

---

# Task 5 — Normality and overall fit {#task-5}

**5a.** Build a **Q-Q plot** of the standardised residuals from `m_log`.

```{r}
# augment(m_log) |> ggplot(aes(sample = .std.resid)) + stat_qq() + stat_qq_line()
```

**5b.** Run Shapiro-Wilk on a sample of 5,000 residuals.

```{r}
set.seed(42)
shapiro.test(sample(resid(m_log), 5000))
```

**Questions:**
- Does the Q-Q plot show heavy tails? What might cause non-normality in wage residuals?
- The Shapiro-Wilk p-value will be very small. Why is this unsurprising and largely irrelevant for inference?

*Your answer:*

**5c.** Compare the posterior predictive check for `m_level` and `m_log`. What does the difference tell you?

```{r}
p1 <- plot(check_predictions(m_level)) + labs(title = "Level-level")
p2 <- plot(check_predictions(m_log))   + labs(title = "Log-level")
p1 + p2
```

*Your answer:*

---

# Task 6 — Full diagnostic check {#task-6}

Run `performance::check_model()` on `m_log`. Match each panel to the task where you built it manually.

```{r}
#| fig-height: 7
performance::check_model(m_log)
```

Based on all evidence, write **3–4 sentences** summarising the diagnostic findings. What are the main concerns, and how did you address (or would you address) each?

*Your answer:*

---

# Task 7 — Omitted variable bias {#task-7}

**7a.** The dataset contains a `female` indicator. Compare a model without and with it:

```{r}
m_no_female   <- lm(log(wage) ~ age + educ,          data = cps)
m_with_female <- lm(log(wage) ~ age + educ + female,  data = cps)

modelsummary(
  list("Without female" = m_no_female, "With female" = m_with_female),
  stars    = TRUE,
  gof_omit = "AIC|BIC|Log|F|RMSE",
  coef_rename = c("age" = "Age", "educ" = "Education", "female" = "Female")
)
```

Does the education coefficient change? In which direction, and why?

*Your answer:*

**7b.** Run `check_model(m_no_female, check = c("linearity", "homogeneity", "qq"))`. Do the residual panels reveal the omission?

```{r}
# your code
```

*Your answer:*

**7c.** Name one variable that is missing from `m_with_female` but that likely confounds the education-wage relationship. Use the direction-of-bias rule to predict whether omitting it overstates or understates the education coefficient.

*Your answer:*

---

# Bonus — RESET test and the quadratic age profile {#bonus}

*Complete if you finish the core tasks early.*

Re-run the RESET test on `m_quad` (the model with `age + I(age^2)`). Does the test result improve relative to `m_log`? What does this suggest about the age-wage relationship?

```{r}
# your code
```

*Your answer:*
